---
title: "Lab 04 - R falls before me!"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summaries and Subscripting
Let's suppose we've collected some data from an experiment and stored them in an object x:

```{r}
x<-c(7.5,8.2,3.1,5.6,8.2,9.3,6.5,7.0,9.3,1.2,14.5,6.2)
```

Some simple summary statistics of these data can be produced:
```{r}
mean(x)
var(x)
sd(x)
summary(x)
```


which should all be self explanatory. 

It may be, however, that we subsequently learn that the first 6 data correspond to measurements made on one machine, and the second six on another machine.

This might suggest summarizing the two sets of data separately, so we would need to extract from
x the two relevant subvectors. This is achieved by subscripting:
```{r}
x[1:6]
x[7:12]
summary(x[1:6])
summary(x[7:12])
```

Other subsets can be created in the obvious way. For example:
```{r}
x[c(2,4,9)]
```

Negative integers can be used to exclude particular elements. For example
```{r}
x[-(1:6)]
```

has the same effect as 
```{r}
x[7:12]
```

### Exercises
1. If x<- c(5,9,2,3,4,6,7,0,8,12,2,9) decide what each of the following is and use R to
check your answers:
(a) x[2]
(b) x[2:4]
(c) x[c(2,3,6)]
(d) x[c(1:5,10:12)]
(e) x[-(10:12)]
2. The data y<-c(33,44,29,16,25,45,33,19,54,22,21,49,11,24,56) contain sales of milk
in gallons for 5 days in three different shops (the frst 3 values are for shops 1,2 and 3 on
Monday, etc.) Produce a statistical summary of the sales for each day of the week and also
for each shop.

### Matrices
Matrices can be created in R in a variety of ways. Perhaps the simplest is to create the columns
and then glue them together with the command cbind. For example,
Matrices can be created in R in a variety of ways. Perhaps the simplest is to create the columns
and then glue them together with the command cbind. For example,
```{r}
x<-c(5,7,9)
y<-c(6,3,4)
z<-cbind(x,y)
z
```


The dimension of a matrix can be checked with the dim command:
```{r}
dim(z)
```

[1] 3 2 i.e., three rows and two columns. There is a similar command, rbind, for building matrices by gluing rows together.  

The functions cbind and rbind can also be applied to matrices themselves (provided the dimensions match) to form larger matrices. For example,
```{r}
rbind(z,z)
```

Matrices can also be built by explicit construction via the function matrix. For example,
```{r}
z<-matrix(c(5,7,9,6,3,4),nrow=3)
```
results in a matrix z identical to z above. Notice that the dimension of the matrix is determined
by the size of the vector and the requirement that the number of rows is 3, as specified by the
argument nrow=3. As an alternative we could have specified the number of columns with the
argument ncol=2 (obviously, it is unnecessary to give both). Notice that the matrix is 'flled up'
column-wise. If instead you wish to fill up row-wise, add the option byrow=T. For example,
```{r}
z<-matrix(c(5,7,9,6,3,4),nr=3,byrow=T)
z
```
Notice that the argument nrow has been abbreviated to nr. Such abbreviations are always
possible for function arguments provided it induces no ambiguity - if in doubt always use the full
argument name.

As usual, R will try to interpret operations on matrices in a natural way. For example, with z
as above, and
```{r}
y<-matrix(c(1,3,0,9,5,-1),nrow=3,byrow=T)
y
```

we obtain
```{r}
y+z
```
and
```{r}
y*z
```

Other useful functions on matrices are to transpose a matrix:
```{r}
z
t(z)
```

As with vectors it is useful to be able to extract sub-components of matrices. In this case, we
may wish to pick out individual elements, rows or columns. As before, the [ ] notation is used to
subscript. The following examples should make things clear:
```{r}
z[1,1]
```

```{r}
z[c(2,3),2]
```

```{r}
z[,2]
```

> 
```{r}
z[1:2,]
```

So, in particular, it is necessary to specify which rows and columns are required, whilst omitting
the integer for either dimension implies that every element in that dimension is selected.

### Exercises


1. Create in R this matrix
```{r, echo = FALSE}
x <-matrix(c(1,7,8,11,-5,3,8,6,3,-9,0,11,14, 14, 14),nrow=3,byrow=T)
x
```
2. Create in R these matrices: 
```{r, echo = FALSE}
x <-matrix(c(1,7,8,11,5,9),nrow=3,byrow=T)
y <-matrix(c(6,8,2,1,1,-7),nrow=3,byrow=T)
x
y
```
3. Calculate the following and check your answers in R:
(a) 2*x
(b) x*x
(c) y%*%x
(d) x%*%y
(e) t(y)
(f ) solve(x)
```{r,echo = FALSE}
x <-matrix(c(1,7,8,11,5,9),nrow=3,byrow=T)
y <-matrix(c(6,8,2,1,1,-7),nrow=3,byrow=T)
2*x
x*x
y%*%x
x%*%y
t(y)
solve(x)
```

4. With x and y as above, calculate the effect of the following subscript operations and check
your answers in R.
(a) x[1,]
(b) x[2,]
(c) x[,2]
(d) y[1,2]
(e) y[,2:3]

