---
title: "Lab 01 - Hello R!"
output: 
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

```{marginfigure}
R is the name of the programming language itself and RStudio is a convenient interface.
```

The main goal of this lab is to introduce you to R and RStudio, which we will be using throughout the course both to learn the statistical concepts discussed in the course and to analyze real data and come to informed conclusions.


### Packages


```{marginfigure}
In R, the fundamental unit of shareable code is the package. 
```

In R, the fundamental unit of shareable code is the package. A package bundles together code, data, documentation, and tests, and is easy to share with others. As of January 2021, there were over 16,000 packages available on the Comprehensive R Archive Network, or CRAN, the public clearing house for R packages. This huge variety of packages is one of the reasons that R is so successful: the chances are that someone has already solved a problem that you’re working on, and you can benefit from their work by downloading their package.


```{marginfigure}
You install them from CRAN with install.packages("x").
You use them in R with library("x").
You get help on them with package?x and help(package = "x").
```

In this lab we will work with two packages: `datasauRus` which contains the dataset, and `tidyverse` which is a collection of packages for doing data analysis in a "tidy" way.

Install these packages by running the following in the console.

```{r eval = FALSE}
install.packages("tidyverse")
install.packages("datasauRus")
```

Now that the necessary packages are installed, you should be able to Knit your document and see the results.

If you'd like to run your code in the Console as well you'll also need to load the packages there.
To do so, run the following in the console.

```{r message=FALSE}
library(tidyverse) 
library(datasauRus)
```

Note that the packages are also loaded with the same commands in your R Markdown document.

## The data

```{marginfigure}
If it's confusing that the data frame is called `datasaurus_dozen` when it contains 13 datasets, you're not alone! Have you heard of a [baker's dozen](https://en.wikipedia.org/wiki/Dozen#Baker's_dozen)?
```

The data frame we will be working with today is called `datasaurus_dozen` and it's in the `datasauRus` package.
Actually, this single data frame contains 13 datasets, designed to show us why data visualisation is important and how summary statistics alone can be misleading.
The different datasets are marked by the `dataset` variable.

To find out more about the dataset, type the following in your Console: `?datasaurus_dozen`.
A question mark before the name of an object will always bring up its help file.
This command must be ran in the Console.

## Exercises

1.  Based on the help file, how many rows and how many columns does the `datasaurus_dozen` file have? What are the variables included in the data frame?  

```{r}
glimpse(datasaurus_dozen)
head(datasaurus_dozen)
tail(datasaurus_dozen)
```

```{marginfigure}
Matejka, Justin, and George Fitzmaurice. "Same stats, different graphs: Generating datasets with varied appearance and identical statistics through simulated annealing." Proceedings of the 2017 CHI Conference on Human Factors in Computing Systems. ACM, 2017.
```

The original Datasaurus (`dino`) was created by Alberto Cairo in [this great blog post](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html).
The other Dozen were generated using simulated annealing and the process is described in the paper *Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics* through Simulated Annealing by Justin Matejka and George Fitzmaurice.
In the paper, the authors simulate a variety of datasets that have the same summary statistics as the Datasaurus but have very different distributions.

2.  Plot `y` vs. `x` for the `dino` dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset.
```{r}
plot(datasaurus_dozen$x, datasaurus_dozen$y)
cor(datasaurus_dozen$x, datasaurus_dozen$y)
```


Below is the code you will need to complete this exercise.
Basically, the answer is already given, but you need to include relevant bits in your Rmd document and successfully knit it and view the results.

Start with the `datasaurus_dozen` and pipe it into the `filter` function to filter for observations where `dataset == "dino"`.
Store the resulting filtered data frame as a new data frame called `dino_data`.

```{r}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")
```

There is a lot going on here, so let's slow down and unpack it a bit.

First, the pipe operator: `%>%`, takes what comes before it and sends it as the first argument to what comes after it.
So here, we're saying `filter` the `datasaurus_dozen` data frame for observations where `dataset == "dino"`.

Second, the assignment operator: `<-`, assigns the name `dino_data` to the filtered data frame.

Next, we need to visualize these data.
We will use the `ggplot` function for this.
Its first argument is the data you're visualizing.
Next we define the `aes`thetic mappings.
In other words, the columns of the data that get mapped to certain aesthetic features of the plot, e.g. the `x` axis will represent the variable called `x` and the `y` axis will represent the variable called `y`.
Then, we add another layer to this plot where we define which `geom`etric shapes we want to use to represent each observation in the data.
In this case we want these to be points, hence `geom_point`.

```{r fig.fullwidth=TRUE}
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

If this seems like a lot, it is.
And you will learn about the philosophy of building data visualizations in layer in detail next week.
For now, follow along with the code that is provided.

For the second part of these exercises, we need to calculate a summary statistic: the correlation coefficient.
Correlation coefficient, often referred to as $r$ in statistics, measures the linear association between two variables.
You will see that some of the pairs of variables we plot do not have a linear relationship between them.
This is exactly why we want to visualize first: visualize to assess the form of the relationship, and calculate $r$ only if relevant.
In this case, calculating a correlation coefficient really doesn't make sense since the relationship between `x` and `y` is definitely not linear -- it's dinosaurial!

But, for illustrative purposes, let's calculate the correlation coefficient between `x` and `y`.

```{marginfigure}
Start with `dino_data` and calculate a summary statistic that we will call `r` as the `cor`relation between `x` and `y`.
```

```{r}
dino_data %>%
  summarize(r = cor(x, y))
```

*This is a good place to pause, commit changes with the commit message "Added answer for Ex 2", and push.*

3.  Plot `y` vs. `x` for the `star` dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset. How does this value compare to the `r` of `dino`?

*This is another good place to pause, commit changes with the commit message "Added answer for Ex 3", and push.*

4.  Plot `y` vs. `x` for the `circle` dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset. How does this value compare to the `r` of `dino`?

*You should pause again, commit changes with the commit message "Added answer for Ex 4", and push.*

```{marginfigure}
Facet by the dataset variable, placing the plots in a 3 column grid, and don't add a legend.
```

5. Finally, let's plot all datasets at once. In order to do this we will make use of facetting.

```{r all-viz, eval=FALSE, fig.fullwidth=TRUE}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")
```

And we can use the `group_by` function to generate all the summary correlation coefficients.

```{r all-r, eval=FALSE}
datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y)) %>%
  print(13)
```

You're done with the data analysis exercises, but we'd like you to do two more things:

```{r fig-resize-global, fig.margin = TRUE, echo = FALSE, fig.width=3}

```

-   **Resize your figures:**

Click on the gear icon in on top of the R Markdown document, and select "Output Options..." in the dropdown menu.
In the pop up dialogue box go to the Figures tab and change the height and width of the figures, and hit OK when done.
Then, knit your document and see how you like the new sizes.
Change and knit again and again until you're happy with the figure sizes.
Note that these values get saved in the YAML.

```{r fig-resize-local, fig.margin = TRUE, echo = FALSE, fig.width=3}

```

You can also use different figure sizes for differen figures.
To do so click on the gear icon within the chunk where you want to make a change.
Changing the figure sizes added new options to these chunks: `fig.width` and `fig.height`.
You can change them by defining different values directly in your R Markdown document as well.

```{r theme-highlight, fig.margin = TRUE, echo = FALSE, fig.width=3}

```

-   **Change the look of your report:**

Once again click on the gear icon in on top of the R Markdown document, and select "Output Options..." in the dropdown menu.
In the General tab of the pop up dialogue box try out different Syntax highlighting and theme options.
Hit OK and knit your document to see how it looks.
Play around with these until you're happy with the look.

<br>

```{marginfigure}
Not sure how to use emojis on your computer? Maybe a teammate can help? Or you can ask your TA as well!
```

*Yay, you're done! Commit all remaining changes, use the commit message "Done with Lab 1!, and push. Before you wrap up the assignment, make sure all documents are updated on your GitHub repo.*
